---
title: "3.2.ExtractCovariatesGEE"
author: "Angela Moreras"
date: "2024-10-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(sf)
library(rgee)

# also need 
# install.packages("geojsonio")

# Issue with ee_extract so I created a modified version. See Issue:
# https://github.com/r-spatial/rgee/issues/367
source("ee_extract_mod.R")

```

Be aware that the library rgee requires several things:
* Google account with Earth Engine activated
* Python >= v3.5 installed 
* EarthEngine Python API (Python package)
* numpy package install 

Follow this easy step by step youtube video to install GEE in R: https://www.youtube.com/watch?v=olNYYynSJfI. Note that the instructions that accompany the rgee package were not working at this time. See the script install_rgee.R for code that accompanies the video.

Instead of using `ee_Initialize()` from rgee, I used the below code to avoid errors in `ee_Initialize` based on the recommendations in [this](https://github.com/r-spatial/rgee/issues/355) issue. 

** That issue also notes that the developer of rgee is potentially not going to maintain it much longer so it might be better to do some of this in python directly. 



#0.Prep

The first time you run the code below you will need to authenticate through your 

```{r}
ee$Authenticate(auth_mode='notebook')
ee$Initialize(project='ee-sarahendicott-eccc')  # <-- EDIT THIS FOR YOUR PROJECT

# Optionally
# make a request to verify you are connected.
ee$String('Hello from the Earth Engine servers!')$getInfo()

#check you have all requirements
#if so all need to be in okay

# This errors for me but not sure if it is a problem:
# ee_check()
```
Note I silence the rgee library for my personal computer to avoid installing new software. 

```{r}
# I think it is already initialized above...
# ee_Initialize()
```
Note that ee_Initialize only works if it does not have any warning<


Then set root path for data on Google drive to the national models and one root the regional models
```{r}
# Not used in this script
# root.drive <- "G:/Shared drives/BAM_NationalModels5"
root <-"."
```

Here we are going to uplaod the spreadsheet that contains all information on the covariates to extract
```{r}
# meth <- read.csv(file.path(root, "Covariate", "RegionalModels_VariableList.csv"),
#                  na.strings=c("NA",""))

# I used the national models list as starting point
meth <- readxl::read_xlsx(file.path(root, "Covariate", "NationalModels_V5_VariableList.xlsx"),
                          na = c("NA",""), sheet = "ExtractionLookup")

# save to a csv where Running and complete columns can be updated as batches are downloaded
write.csv(meth,
  file = file.path(root, "Covariate", "Model_VariableList.csv"),
  row.names = FALSE
)

meth <- read.csv(file.path(root, "Covariate", "Model_VariableList.csv"),
                 na.strings=c("NA",""))

```

Set the proper projection. 
Note that this project uses CONUS Albers equal-area conic (EPSG: 5072) 
```{r}
projection.trans<-5072
projection.st_as_sf<-4326
```

#1. Data prep

Load data, leaving only the visit object (created in 1.3.DataWrangling)
```{r}
# load(file.path(root, "1_RM_data_clean.R")) 
# rm(bird)

# try using data I downloaded and processed separately
analysis_data_package <- readRDS(file.path(
  "..", "ROFBirds",
  "analysis/data/derived_data/INLA_data/analysis_data_package.rds"
))

visit <- analysis_data_package$all_surveys %>% 
  sf::st_drop_geometry() %>% 
  # rename to fit WT format
  dplyr::mutate(project_id = Project_Name, location_id = survey_ID, 
         latitude = Latitude, longitude = Longitude, 
         year = lubridate::year(Date_Time), .keep = "none")

```

##1.2. Create sf object of just location*year for annual layers
```{r}
loc.yr <- visit %>% 
  dplyr::select(project_id, location_id, latitude, longitude, year) %>%
  dplyr::mutate(year=as.integer(year))%>%
  unique() %>% 
  dplyr::mutate(id=paste(project_id, location_id, latitude, longitude, year, sep="_")) %>% 
  st_as_sf(coords=c("longitude", "latitude"), crs=projection.st_as_sf, remove=FALSE) %>% 
  st_transform(crs=projection.trans)
```

##EXTRA. Test sample dataset of loc.yr
```{r}
#run this lines to use a test sample
#set.seed(1234)
#loc.n <- loc.yr %>%
 # sample_n(1000)

#run this line to use all data 
loc.n <- loc.yr
```

##1.3. Buffer location objects
200 m radius for local extent and 2 km radius to landscape extent.
Please be sure that the local.buffer and the neighbour.buff matched the ones in the spreadsheet. 

```{r}
local.buffer<-200
neighbour.buffer<-2000

loc.buff <- st_buffer(loc.n, local.buffer)
loc.buff2 <- st_buffer(loc.n, neighbour.buffer)
```


#2.Extract Temporally Static

##2.1 Get list of static layers to run
```{r}
meth.gee <- dplyr::filter(meth, Source=="Google Earth Engine", Use==1, TemporalResolution=="static")
```

##2.2. Extract study area and its extent
One can crop the GEE data into the data study extent by using a shapefile
```{r}
# study_area<-read_sf(file.path(root, "Study_area/BCR81214_Can/BCR81214_Can.shp"))

study_area <- summarise(analysis_data_package$ONGrid) %>% st_transform(crs = 4326)

study_area_bbox<-sf::st_bbox(study_area)
study_area_bounds<-ee$Geometry$Rectangle(
  c(
    west= study_area_bbox[["xmin"]],
    south= study_area_bbox[["ymin"]],
    east= study_area_bbox[["xmax"]],
    north= study_area_bbox[["ymax"]]
  ),
  geodetic=T,
  proj= "EPSG:4326"
)
```

example
```{r}

loc.buff.small<-loc.buff %>% slice(1:20)
gee_data <-ee$Image(meth.gee$Link[3])
# Not working because SpatialResolution is a character
# gee_data1<-ee_extract(gee_data, loc.buff.small, fun = ee$Reducer$mean(),
#   scale = as.integer(meth.gee$SpatialResolution[3]))

# Since National file says to use native I am extracting that from the object and using it. 
nominal_scale <- gee_data$projection()$nominalScale()$getInfo()

# set scale to default for now
gee_data1 <- ee_extract(gee_data, loc.buff.small, fun = ee$Reducer$mean(), scale = nominal_scale)
```

##2.3. Create a plain dataframe 

```{r}
loc.gee.static <-data.frame(loc.n) %>%
   dplyr::select(-geometry)
```

##2.4. Make method loop
```{r}
for(i in 1:nrow(meth.gee)){
  print(i)
  if(meth.gee$GEEtype[i]=="image"){
    if(!is.na(meth.gee$GEEBand[i]))
      img.i <- ee$Image(meth.gee$Link[i])$
      select(meth.gee$GEEBand[i])
    else img.i <- ee$Image(meth.gee$Link[i]) 
  }
  if(meth.gee$GEEtype[i]=="imagecollection"){
    img.i <- ee$ImageCollection(meth.gee$Link[i])$
      select(meth.gee$GEEBand[i])$
      toBands()
  }
  
  loc.gee.i <- switch(as.character(meth.gee$RadiusExtent[i]), 
                      "200"=loc.buff, #buffer width!
                      "2000"=loc.buff2, #buffer width!
                      "NA"=loc.n) 
  
  n = 1000 #number of calls per query/ 
            #GEE only accepts 5000 at time, but payload are restricted then 1000 was chosen
  loc.gee.i<-loc.gee.i%>%
    select(id)%>%
    group_by(row_number() %/% n) %>% group_map(~ .x)
    
  
  gee.data.static<-data.frame()
  
  for (j in 1:length(loc.gee.i)){
    print("check mean or cv")  
    if(meth.gee$RadiusFunction[i]=="mean"){
      gee.data.static.extract <- ee_extract(img.i, loc.gee.i[[j]], 
                              fun = ee$Reducer$mean(),
                              scale = as.integer(meth.gee$GEEScale[i]))
    }
    
    if(meth.gee$RadiusFunction[i]=="cv"){
      gee.data.static.extract <- ee_extract(img.i, loc.gee.i[[j]],
                              fun= ee$Reducer$stdDev(),
                              scale =as.integer(meth.gee$GEEScale[i]))
    }
    
    if(ncol(gee.data.static.extract)<2){ #Corrects for extraction when all data NA
      gee.data.static.extract<-  gee.data.static.extract %>%
      mutate(namecol=NA)
      names(gee.data.static.extract)[2]<-meth.gee$GEEBand[i]
    }
    
    gee.data.static<-rbind(gee.data.static, gee.data.static.extract)
    print(paste0(j, " of ", length(loc.gee.i)))
  }
  print("left join")
  names(gee.data.static)[2]<-meth.gee$Label[i]
  loc.gee.static <- left_join(loc.gee.static, gee.data.static)  
    
  print(paste0("Finished ", i, " of ", nrow(meth.gee)))
  
  meth$Complete[which(meth$Label == meth.gee$Label[i])] <- 1
  meth$Running[which(meth$Label == meth.gee$Label[i])] <- 0
}

#Zerofill
zerocols <- meth.gee %>% 
  dplyr::filter(Zerofill==1)
loc.gee.static <- loc.gee.static%>%
  mutate_at(vars(zerocols$Label), ~replace(., is.na(.), 0))

``` 

##2.5. Save 
```{r}
write.csv(loc.gee.static, file=file.path(root, "Covariate", "03_RM_data_covariates_GEE-static.csv"), row.names = FALSE)

write.csv(meth,file=file.path(root, "Covariate", "Model_VariableList.csv"),
          row.names = FALSE)
```

#3. Extract Temporally matched

##3.1. Get list of static layers to run
```{r}
meth.gee <- dplyr::filter(meth, Source=="Google Earth Engine", TemporalResolution=="match")
```

##3.2. Plain dataframe for joining to output
````{r}
loc.gee.match <- data.frame(loc.n) %>% 
  dplyr::select(-geometry)
```


##3.3. Set up to loop
```{r}

#TODO: no values being extracted for i = 9 and j = 20 and h = 2
for(i in 9:nrow(meth.gee)){
  
  #Identify years of imagery
  years.gee <- seq(meth.gee$GEEYearMin[i], meth.gee$GEEYearMax[i]) 
    #considers having all years 
  
  #Match year of data to year of data
  dt = data.table::data.table(year=years.gee, val=years.gee)
  data.table::setattr(dt, "sorted", "year")
  data.table::setkey(dt, year)
  
  loc.n.i <- switch(as.character(meth.gee$RadiusExtent[i]), 
                    "200"= loc.buff, #buffer width!
                    "2000"=loc.buff2, #buffer width!
                    "NA"=loc.n) 
  
  loc.n.i$yearrd <- dt[J(loc.n.i$year), roll = "nearest"]$val
  
  #Set up to loop through years
  loc.j <- data.frame()
  for(j in 1:length(years.gee)){
    
    loc.n.yr <- dplyr::filter(loc.n.i, yearrd==years.gee[j]) %>%
      select (id)

    if(nrow(loc.n.yr) > 0){
      
      #Set start & end date for image filtering---
      start.k <- paste0(years.gee[j]+meth.gee$YearMatch[i], "-",
                          meth.gee$GEEMonthMin[i], "-01")
        
      if(meth.gee$GEEMonthMax[i] > meth.gee$GEEMonthMin[i]){
        end.k <- paste0(years.gee[j]+meth.gee$YearMatch[i], "-",
                          meth.gee$GEEMonthMax[i], "-28")
      }
      if(meth.gee$GEEMonthMax[i] < meth.gee$GEEMonthMin[i]){
          end.k <- paste0(years.gee[j], "-", meth.gee$GEEMonthMax[i], "-28")
      }
        
      #Get the image
      img.i <- ee$ImageCollection(meth.gee$Link[i])$
        filter(ee$Filter$date(start.k, end.k))$select(meth.gee$GEEBand[i])$mean()
        
      #Extract
      if (nrow(loc.n.yr)>1000) { #5000 entries is the limit for google earth engine
                                #Some maps exceed the payload size, therefore we selected 1000
        n = 1000 
        split <- loc.n.yr %>% group_by(row_number() %/% n) %>% group_map(~ .x)
      }
      
      if (nrow(loc.n.yr)<=1000){
        split <- list (loc.n.yr) 
      }
      
      gee.data<-data.frame()
      for (h in 1:length(split)){
        if(meth.gee$Extraction[i]=="point"){
          gee.ext.data <- ee_extract(img.i, split[[h]],
                               scale =as.integer(meth.gee$GEEScale[i]))
        }
      
        if(meth.gee$Extraction[i]=="radius"){
          
          if(meth.gee$RadiusFunction[i]=="mean"){
            gee.ext.data <- ee_extract(img.i, split[[h]],
                                  fun= ee$Reducer$mean(),
                                  scale =as.integer(meth.gee$GEEScale[i])) 
         }
          
          if(meth.gee$RadiusFunction[i]=="mode"){
            gee.ext.data <- ee_extract(img.i, split[[h]],
                                  fun= ee$Reducer$mode(),
                                  scale =as.integer(meth.gee$GEEScale[i])) 
          }
        } 
        
        if(ncol(gee.ext.data)<2){ #Corrects for extraction when all data NA
          gee.ext.data<-  gee.ext.data %>%
          mutate(namecol=NA)
          names(gee.ext.data)[2]<-meth.gee$GEEBand[i]
        }  
        gee.data <- rbind(gee.data, gee.ext.data)     
      }
      
    #Collapse data across years
    names(gee.data)[2]<-meth.gee$Label[i]
    loc.j<-rbind(loc.j, gee.data)
    }
    
    print(paste0("Finished year ", j, " of ", length(years.gee)))
    
  }
  loc.gee.match <- left_join(loc.gee.match, loc.j) 
  print(paste0("Finished ", i, " of ", nrow(meth.gee)))
  
  meth$Complete[which(meth$Label == meth.gee$Label[i])] <- 1
  meth$Run[which(meth$Label == meth.gee$Label[i])] <- 0
}
```

##3.4. Save
```{r}
write.csv(loc.gee.match, file=file.path(root, "Covariate", "03_RM_data_covariates_GEE-match.csv"), 
          row.names = FALSE)

write.csv(meth,file=file.path(root, "Covariate", "Model_VariableList.csv"),
          row.names = FALSE)
```
